---
title: 'Wk6: Genome Informatics'
author: "Yiyu"
date: "2025-02-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1. Proportion of G/G population

Downloaded a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39895035-39895162;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel

Here we read the CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl) *100
```

#Section 4: Population Scale Analysis [HOMEWORK]
One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```
> Q13. Determine sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
table(expr$geno)
```
Sample size for A/A is 108, A/G is 233, G/G is 121.

```{r}
library(dplyr)
median <- expr |>
  group_by(geno) |>
  summarize(median_expression = median(exp))
print(median)
```
Can use **dplyr** package to print median expression for each genotype using the `group_by` function.

```{r}
library(ggplot2)
```

>Q14. Generate a boxplot wiht a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP affect the expression of ORMDL3?

Let's make a boxplot with our data
```{r}
ggplot(expr) + aes(geno, exp, fill = geno) +
  geom_boxplot(notch = TRUE)
```
Yes, the box plot indicates that having G/G correlates to reduced expression of ORMDL3 gene, while A/A correlates to higher expression.
