---
title: "AlphaFold analysis"
format: pdf
---

Here we demonstrate how to analyze and make sense of models from AlphaFold. We begin by reading all the model PDB files....

```{r}
library(bio3d)
```

PDB file names of my models
```{r}
files <- list.files("hiv_monomer_94b5b_0/", 
           pattern = ".pdb",
           full.names = T)
```

# Align and superimpose

```{r}
pdbs <- pdbaln(files, fit=TRUE, exefile="msa")
```
RMSD analysis

```{r}
rd <- rmsd(pdbs)
mean(rd)
```
```{r}
source("https://tinyurl.com/newviewngl")
library(NGLVieweR)
```

```{r}
#view.pdbs(pdbs)
```

#PCA
```{r}
pc <- pca(pdbs)
plot(pc)
```

#Residue conservation from alignment file

AlphaFold writes out the MSA it calculated and used for structure prediction to a A3M format file that we can read into R for further analysis:


```{r}
aln_file <- list.files(path="hiv_monomer_94b5b_0/",
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file
```

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```
```{r}
dim(aln$ali)
```

We can score residue conservation:

```{r}
sim <- conserv(aln)
```

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```

Plot the conservation along the sequence/structure
```{r}
plot(sim, type = "h")
```
Let's look at these conversved positions in the structure:
```{r}
pdb <- read.pdb( files[1] )

#view.pdb(pdb,  backgroundColor ="pink",
 #        highlight = atom.select(pdb,resno=25:28),
  #       highlight.style = "spacefill")
```

