---
title: "Class18: Pertussis mini project"
author: "Yiyu"
format: pdf
toc: TRUE
---

Pertussis (aka whooping cough) is a serious lung infection caused by the bacteria *B. Pertussis.*

The CDC tracks Pertussis case numbers and we can fin dthis data here: 
http://tinyurl.com/pertussiscdc

We can "scrape" this data using the **datapasta** package. Download updated package from github and install in R console.

```{r, include=FALSE}
cdc <- data.frame(
  year = c(1922L,1923L,1924L,1925L,
           1926L,1927L,1928L,1929L,1930L,1931L,
           1932L,1933L,1934L,1935L,1936L,
           1937L,1938L,1939L,1940L,1941L,1942L,
           1943L,1944L,1945L,1946L,1947L,
           1948L,1949L,1950L,1951L,1952L,
           1953L,1954L,1955L,1956L,1957L,1958L,
           1959L,1960L,1961L,1962L,1963L,
           1964L,1965L,1966L,1967L,1968L,1969L,
           1970L,1971L,1972L,1973L,1974L,
           1975L,1976L,1977L,1978L,1979L,1980L,
           1981L,1982L,1983L,1984L,1985L,
           1986L,1987L,1988L,1989L,1990L,
           1991L,1992L,1993L,1994L,1995L,1996L,
           1997L,1998L,1999L,2000L,2001L,
           2002L,2003L,2004L,2005L,2006L,2007L,
           2008L,2009L,2010L,2011L,2012L,
           2013L,2014L,2015L,2016L,2017L,2018L,
           2019L,2020L,2021L,2022L, 2024L),
  cases = c(107473,164191,165418,152003,
                                   202210,181411,161799,197371,
                                   166914,172559,215343,179135,265269,
                                   180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,
                                   120718,68687,45030,37129,60886,
                                   62786,31732,28295,32148,40005,
                                   14809,11468,17749,17135,13005,6799,
                                   7717,9718,4810,3285,4249,3036,
                                   3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,
                                   3589,4195,2823,3450,4157,4570,
                                   2719,4083,6586,4617,5137,7796,6564,
                                   7405,7298,7867,7580,9771,11647,
                                   25827,25616,15632,10454,13278,
                                   16858,27550,18719,48277,28639,32971,
                                   20762,17972,18975,15609,18617,
                                   6124,2116,3044, 35435)
)
```

```{r}
head(cdc)
```

> Q1. Make a plot of pertussis cases per year using ggplot.

```{r}
library(ggplot2)
ggplot(cdc) +
  aes(year, cases) +
  labs(title = "pertussis cases per year") +
  geom_line()
```

> Q2. Let's add key milestones of the DTP (wP) vaccine roll out in 1942 and switch to the new aP vaccine in 1996. We can use `geom_vline()` to add vertical line for these. Booster shots started in 2004, while covid quarantine began in 2020.

```{r}
ggplot(cdc) +
  aes(year, cases) +
  labs(title = "pertussis cases per year") +
  geom_line() +
  geom_vline(xintercept = 1942, col = "blue") +
  geom_vline(xintercept = 1996, col = "red") + 
  geom_vline(xintercept = 2004, col = "violet") +
  geom_vline(xintercept = 2020, col= "orange")
```

There were high case numbers pre 1946 (before the wP vaccine), then relatively rapid decrease in cases post wP vaccine. The case numbers remained low throughout the 1970s until 2004 when the first widespread outbreak occurred again. 

In 1996 there was a switch from the wP vaccine to the aP vaccine, and soon after in 2004 there was an increase in Pertussis cases in adolescents. 

Mounting evidence indicates that the aP vaccine induced immunity wanes faster than the older wP vaccine.

## Computational Models of Immunity Pertussis Boost

One of the main goals of this project is to determine what is different in the immune repsonse between wP and aP primed individuals.

Using the booster vaccine as a proxy for ineffection


All data is available here https://www.cmi-pb.org/ in JSON format. We can use the **jsonlite** package to read this data into R.

```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE) 

head(subject)
```

> Q. How many individuals "subjects"

```{r}
nrow(subject)
```

> Q. How many aP and wP subjects are there?

```{r}
table(subject$infancy_vac)
```

> Q. Male/Female numbers

```{r}
table(subject$biological_sex)
```

> Q. Breakdown of biological sex and race?

```{r}
table(subject$race, subject$biological_sex)
```

> Q. Does this look to be representitive of the US population at large?

NO. This is not representative of the US population at large, it is biased.

Let's read some more CMI-PB data

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", 
                      simplifyVector = TRUE) 
titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer",
                   simplifyVector = TRUE) 
```

```{r}
head(specimen)
```
```{r}
head(titer)
```
To use this data we need to "join" the various tables to find all the information I need to know about a particular measurement.

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
dim(meta)
head(meta)
```

Now we can join `meta` with `titer` data.

```{r}
ab_data <- inner_join(meta, titer)
head(ab_data)
```

> Q. How many different antibody isotypes are we measuring?

```{r}
table(ab_data$isotype)
```

> Q. How many antigens?

```{r}
table(ab_data$antigen)
```
> Q. Let's look at antigen levels over the whole dataset?

```{r}
dim(ab_data)
```

Let's look at a boxplot of antigen levels 
```{r}
ggplot(ab_data) +
  aes(MFI, antigen) +
  geom_boxplot()
```

> Q. Break this plot down by aP or wP.

We can use coloring `col=` by `infancy_vac`.

```{r}
ggplot(ab_data) +
  aes(MFI, antigen, col = infancy_vac) +
  geom_boxplot()
```

Or we can `facet` the plot by `~infancy_vac`

```{r}
ggplot(ab_data) +
  aes(MFI, antigen) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

Let's try using MFI_normalized instead of MFI

```{r}
ggplot(ab_data) +
  aes(MFI_normalised, antigen, col = infancy_vac) +
  geom_boxplot()
```

Let's focus on just IgG

```{r}
igg <- ab_data |>
  filter(isotype == "IgG")

ggplot(igg) +
  aes(MFI_normalised, antigen, col = infancy_vac) +
  geom_boxplot()
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col = infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit)
```

Let's focus on PT (pertussis toxin) and igg levels over time. To do this we'll filter to focus on one antigen "PT" and IgG levels in 2021 dataset.

```{r}
pt_igg_2021 <- ab_data |>
      filter(isotype == "IgG", antigen == "PT", dataset == "2021_dataset")
```

A plot of `actual_day_relative_to_boost_` vs `MFI_normalised`
```{r}
ggplot(pt_igg_2021) + 
  aes(actual_day_relative_to_boost, MFI_normalised, col=infancy_vac, 
      group = subject_id) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 14) +
  geom_vline(xintercept = 0) +
  labs(title = "2021 IgG PT levels",
       xlab = "Days relative to Boost",
       ylab = "MFI (normalized)",
       subtitle = "Vertical lines indicate day 0 (pre-boost) and 14 
       (apparent peak levels)") 

```

```{r}
library(lubridate)
```

```{r}
today () - mdy("04-23-2001")
```

```{r}
time_length( today() - mdy("04-23-2001"), "years")
```

```{r}
subject$age <- time_length(today() - ymd(subject$year_of_birth), "years")
```

