---
title: "Class 5: Data Viz with ggplot"
author: "Yiyu"
format: pdf
---
# Graphs and plots in R

R has tons of different graphics systems. These include **"base R"** (e.g. the `plot()` function) and add on packages like **ggplot2**.

Let's start with plotting a simple data set in "base R" and then ggplot2 to see how they differ.

```{r}
head(cars)
```

To plot this in "base R", I just use `plot()`

```{r}
plot(cars)
```
First to use ggplot2, I need to install the package. For this I use the `install.packages()` function

> I don't want to run `install.packages()` in my quarto document as this would re-install the package every time I render the document.

The main function in the ggplot2 package is `ggplot()`, before I can use this function I need to load the package with a `library()` call.
```{r}
library(ggplot2)
ggplot(cars)
```
Blank because it's missing layers

There are at least 3 things that every ggplot needs: 

- the **data** (the data I want to plot)
- the **aes**thetics (how the data maps to my plot) 
- the **geoms** or geometries (the type of plot)

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point()
```

Can add even more layers

```{r}
p <- ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() + 
  geom_smooth(method="lm")
```
can save it to p, then just add on to p

```{r}
p
```

```{r}
p + labs(title="My very nice GGPLOT", x= "Speed(MPH)") + 
  theme_bw()
```

##RNSeq plot with more aes() values

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)

```
> Q. Use the nrow() function to find out how many genes are in this dataset. What is your answer? 

> Q. Use the colnames() function and the ncol() function on the genes data frame to find out what the column names are (we will need these later) and how many columns there are. How many columns did you find? 

> Q. Use the table() function on the State column of this data.frame to find out how many ‘up’ regulated genes there are. What is your answer? 

> Q. Using your values above and 2 significant figures. What fraction of total genes is up-regulated in this dataset? 

```{r}
nrow(genes)
colnames(genes)
ncol(genes)
table(genes$State)
table(genes$State) / nrow(genes) * 100
```

```{r}
ggplot(genes) +aes(x=Condition1, y=Condition2)+geom_point()
```

```{r}
ggplot(genes) + aes(x=Condition1, y=Condition2, col=State)+ geom_point() +
  scale_color_manual(values=c("blue","darkgray","red")) + 
  labs(title= "Expression changes upon drug treatment", subtitle= "RNASeq Data", 
       x= "Control (no drug)", y= "Drug treatment")
```

##Gapminder dataset plots

We can get exposure to setting more `aes()` parameters with datasets that include more columns of useful data. For example the **gapminder** dataset on GDP and life expectancy for different countries over time.

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

>Have a look at the first 6 rows

```{r}
head(gapminder)
```
>How many rows does this dataset have?

```{r}
nrow(gapminder)
```
>How many unqiue contries are there in this dataset?

```{r}
length(unique(gapminder$country))
```

>What years does this dataset cover?

```{r}
unique(gapminder$year)
```

Key functions that will be useful in our R journey include: 
- `nrow()`
- `ncol()`
- `length()`
- `unique()`
- `table()`

```{r}
ggplot(gapminder) + aes(x=gdpPercap, y=lifeExp, col=continent) +
  geom_point(alpha=0.3) + 
  facet_wrap(~continent)
```

```{r}
library(patchwork)
# Setup some example plots 
p1 <- ggplot(mtcars) + geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) + geom_boxplot(aes(gear, disp, group = gear))
p3 <- ggplot(mtcars) + geom_smooth(aes(disp, qsec))
p4 <- ggplot(mtcars) + geom_bar(aes(carb)) 

# Use patchwork to combine them here:
(p1 | p2 | p3) /
      (p4 | p)
```


